---
title: 'Practical Machine Learning: Prediction Assignment Writeup'
author: "I. Chivite"
date: "October/2017"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

The goal of this project is to predict the manner in which six subjects did a particular exercise using  [data](http://groupware.les.inf.puc-rio.br/har) from accelerometers on the belt, forearm, arm, and dumbbell. This data was made available by the [group of research and development of groupware technologies](http://groupware.les.inf.puc-rio.br/) which is specialized, among other things, in *Human Activity Recognition*.    


## Introduction 

Participants were asked to perform one set of 10 repetitions of the *Unilateral Dumbbell Biceps Curl* in five different ways: 

* **Class A:** Exactly according to the specification
* **Class B:** Throwing the elbows to the front
* **Class C:** Lifting the dumbbell only halfway 
* **Class D:** Lowering the dumbbell only halfway
* **Class E:** Throwing the hips to the front 

Class A is the right way, while the other 4 classes correspond to common mistakes. A full description of the study and methodology is available at [*Velloso, et al*](http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201)

## Feature extraction and selection

![Sensor Configuration. Figure from [*Velloso, et al*](http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201)](werables.jpeg)

* Readings from sensors on the belt, forearm, arm, and dumbbell were collected on each of the six subjects in overlapping time windows of 2.5 s.   

* The Euler angles (roll, pitch and yaw) were specified together with other derived features.    

* Raw accelerometer, gyroscope and magnetometer readings were also recorded.  

* For the Euler angles of each of the four sensors, derived features were obtained. They are: mean, variance, standard deviation, max and min.   

* The set of criteria for the choice of these features and derived quantities, a total seventeen variables, is described in the original publication <sup>[1](#myfootnote1)</sup>.   

#### Features selected: (mean, var, max and min per window)  

**Belt**

|    |   Feature   |   Column Name   |   Associated variables   |   
|----:|------------:|---------------:|--------------------------:|
|1 | mean roll | avg_roll_belt|   roll_belt   |    
|2 | variance of the roll  |   var_roll_belt  |   roll_belt   |   
|3 |  maximum of the accelerometer vector|    |accel_belt_x,  accel_belt_y,  accel_belt_z|     
|4   |  range of the accelerometer vector |   |accel_belt_x,  accel_belt_y,  accel_belt_z|     
|5 |   variance of the accelerometer vector|   |   accel_belt_x,  accel_belt_y,  accel_belt_z|   
|6  |   variance of the gyro|   |   gyros_belt_x,  gyros_belt_y, gyros_belt_z   |   
|7  |   variance of the magnetometer|   | magnet_belt_x,	magnet_belt_y  |   
  

**Arm**   

|    |   Feature   |   Column Name   |   Associated variables   |   
|----:|------------:|---------------:|--------------------------:|
|8 | variance of the accelerometer vector: | var_accel_arm| accel_arm_x,	accel_arm_y,	accel_arm_z     |    
|9 | maximum of the magnetometer |     |  magnet_arm_x,	magnet_arm_y,	magnet_arm_z   |
|10 |   minimum of the magnetometer|    |magnet_arm_x,	magnet_arm_y,	magnet_arm_z|     


**Dumbbell**

|    |   Feature   |   Column Name   |   Associated variables   |   
|----:|------------:|---------------:|--------------------------:|
|11 | maximum of the acceleration | total_accel_dumbbell |   accel_dumbbell_x, accel_dumbbell_y, accel_dumbbell_z   |    
|12 | variance of the gyro  |    |   gyros_dumbbell_x,	gyros_dumbbell_y,	gyros_dumbbell_z   |   
|13 |  maximum of magnetometer|    |magnet_dumbbell_x,	magnet_dumbbell_y,	magnet_dumbbell_z|     
|14   |  minimum of the magnetometer |   |magnet_dumbbell_x	magnet_dumbbell_y	magnet_dumbbell_z|    

**Glove** 


|    |   Feature   |   Column Name   |   Associated variables   |   
|----:|------------:|---------------:|--------------------------:|
|15 | sum of the pitch |  |   pitch_forearm, avg_pitch_forearm   |    
|16 | maximum of the gyro  |    |   gyros_forearm_x,	gyros_forearm_y,	gyros_forearm_z   |   
|17 |  minimum of the gyro|    |gyros_forearm_x,	gyros_forearm_y,	gyros_forearm_z|   

The derived features were calculated using the Euclidean magnitude of the associated vector:

$$
F=\left<\sqrt{F_x^2+F_y^2+F_z^2}\right>
$$

Where the average is taken over the corresponding time window (num_window). After this, the max, min, var or just the magnitude are taken.   

## Processing of the Data

Load the necessary libraries:

```{r load libraries, cache=TRUE}
library(caret)
```

Read the raw data:

```{r read the training data}
dataRaw <- read.csv(file="pml-training.csv",head=TRUE)
#names(dataRaw)
```

Extract the ready available features: 

```{r subset for relevant features, cache=TRUE}
relFeatures<-c("num_window","new_window","avg_roll_belt","var_roll_belt","var_accel_arm","total_accel_dumbbell","classe")
relvDataRaw <- dataRaw[,relFeatures]
relvData<-subset(relvDataRaw,new_window=="yes")
```

Proceed to find the derived features:

```{r Derived features, cache=TRUE}

VariablesNames<-c("num_window","new_window", "accel_belt_x","accel_belt_y","accel_belt_z","gyros_belt_x","gyros_belt_y","gyros_belt_z","magnet_belt_x","magnet_belt_y", "magnet_arm_x","magnet_arm_y","magnet_arm_z", "gyros_dumbbell_x","gyros_dumbbell_y","gyros_dumbbell_z","magnet_dumbbell_x","magnet_dumbbell_y","magnet_dumbbell_z","pitch_forearm","gyros_forearm_x","gyros_forearm_y","gyros_forearm_z")

Data<-dataRaw[,VariablesNames]

# delete rows with calculated averages, we have already taken them

Data<-Data[-(Data$new_window=="yes"),]

Data$mag_acc_belt<-sqrt((Data$accel_belt_x)^2+(Data$accel_belt_y)^2+(Data$accel_belt_z)^2)

Data$mag_gyro<-sqrt((Data$gyros_belt_x)^2+(Data$gyros_belt_y)^2+(Data$gyros_belt_z)^2)

Data$mag_magnet<-sqrt((Data$magnet_belt_x)^2+(Data$magnet_belt_y)^2)

max_acc_belt<-aggregate(mag_acc_belt~num_window,data=Data,max)

colnames(max_acc_belt)[2]<-"max_acc_belt"

min_acc_belt<-aggregate(mag_acc_belt~num_window,data=Data,min)

colnames(min_acc_belt)[2]<-"min_acc_belt"

var_acc_belt<-aggregate(mag_acc_belt~num_window,data=Data,var)

colnames(var_acc_belt)[2]<-"var_acc_belt"

var_gyro_belt<-aggregate(mag_gyro~num_window,data=Data,var)

colnames(var_gyro_belt)[2]<-"var_gyro_belt"

var_magnet_belt<-aggregate(mag_magnet~num_window,data=Data,var)

colnames(var_magnet_belt)[2]<-"var_magnet_belt"

# Merge the data sets for belt 

relvData<-merge(relvData,max_acc_belt,by="num_window")

relvData<-merge(relvData,min_acc_belt,by="num_window")

relvData<-merge(relvData,var_acc_belt,by="num_window")

relvData<-merge(relvData,var_gyro_belt,by="num_window")

relvData<-merge(relvData,var_magnet_belt,by="num_window")

```
Now for the **arm** data

```{r arm data, cache=TRUE}

Data$mag_accel_arm<-sqrt((Data$magnet_arm_x)^2+(Data$magnet_arm_y)^2+(Data$magnet_arm_z)^2)

max_acc_arm<-aggregate(mag_accel_arm~num_window,data=Data,max)

colnames(max_acc_arm)[2]<-"max_acc_arm"

min_acc_arm<-aggregate(mag_accel_arm~num_window,data=Data,min)

colnames(min_acc_arm)[2]<-"min_acc_arm"

# Merge Arm data

relvData<-merge(relvData,max_acc_arm,by="num_window")

relvData<-merge(relvData,min_acc_arm,by="num_window")

```

 Proceed with the **dumbbell** data:

```{r dumbell data, cache=TRUE}

Data$mag_gyro_dumbbell<-sqrt((Data$gyros_dumbbell_x)^2+(Data$gyros_dumbbell_y)^2+(Data$gyros_dumbbell_z)^2)

gyro_dumbbell_var<-aggregate(mag_gyro_dumbbell~num_window,data=Data,var)

colnames(gyro_dumbbell_var)[2]<-"gyro_dumbbell_var"

Data$mag_magnet_dumbbell<-sqrt((Data$magnet_dumbbell_x)^2+(Data$magnet_dumbbell_y)^2+(Data$magnet_dumbbell_z)^2)

max_magnet_dumbbell<-aggregate(mag_magnet_dumbbell~num_window,data=Data,max)

colnames(max_magnet_dumbbell)[2]<-"max_magnet_dumbbell"

min_magnet_dumbbell<-aggregate(mag_magnet_dumbbell~num_window,data=Data,min)

colnames(min_magnet_dumbbell)[2]<-"min_magnet_dumbbell"

# Merge dumbbell data

relvData<-merge(relvData,gyro_dumbbell_var,by="num_window")

relvData<-merge(relvData,max_magnet_dumbbell,by="num_window")

relvData<-merge(relvData,min_magnet_dumbbell,by="num_window")

```
And finally the **glove** data

```{r Glove data, cache=TRUE}
Data$mag_gyros_forearm<-sqrt((Data$gyros_forearm_x)^2+(Data$gyros_forearm_y)^2+(Data$gyros_forearm_z)^2)

max_gyros_forearm<-aggregate(mag_gyros_forearm~num_window,data=Data,max)

colnames(max_gyros_forearm)[2]<-"max_gyros_forearm"

min_gyros_forearm<-aggregate(mag_gyros_forearm~num_window,data=Data,min)

colnames(min_gyros_forearm)[2]<-"min_gyros_forearm"

# Merge glove data

relvData<-merge(relvData,max_gyros_forearm,by="num_window")

relvData<-merge(relvData,min_gyros_forearm,by="num_window")

```
Now some house cleaning,

```{r remove unnecessary columns}

drops<-c("new_window")
relvData<-relvData[,!(names(relvData)%in% drops)]
#names(relvData)
```

## Model and Model Prediction

Random Forest algorithm was chosen for the predictions. There is a collection of weak independent
features (variables) that need to be aggregated. Hence the decision of boosting is natural<sup>[1](#myfootnote1)</sup>.

For cross validation the training set was divided into two different sets: training and tr_testing. The **training** set was used to produce the model and the **tr_testing** set was used for estimation of sample rates.

```{r dividing the training set, cache=TRUE}
set.seed(12345)#
inTrain <- createDataPartition(y=relvData$classe,p=0.8, list=FALSE)
training <- relvData[inTrain,]
tr_testing <- relvData[-inTrain,]
dim(training)
dim(tr_testing)
training<-na.omit(training)
tr_testing<-na.omit(tr_testing)
dim(training)
dim(tr_testing)
```
Proceed to model

```{r random forest, cache=TRUE}
modelRF <- train(classe~.-(num_window),data=training,method="rf",prox=TRUE)
modelRF
```
Estimate the sample error rates:

```{r prediction, cache=TRUE}

predRF<-predict(modelRF,newdata=tr_testing) 
confusionMatrix(predRF,tr_testing$classe)

```

## Predictions for the Test Data:

The test data consists of 20 different cases taken from individual time windows. Since derived features were used to make predictions, it is necessary some pre-processing. 

The derived features for the test set were calculated using the information available for the corresponding window following the same definitions used for the training set.


```{r read the test data,  include=FALSE, echo=FALSE}

dataRaw_2 <- read.csv(file="pml-testing.csv",head=TRUE)

testing<-as.data.frame(dataRaw_2$num_window)

colnames(testing)[1]<-"num_window"

testing<-merge(testing,relvData,by="num_window")

testing

difference<-setdiff(dataRaw_2$num_window, testing$num_window)

drops<-c("classe")
testing<-testing[,!(names(relvData)%in% drops)]
```

```{r finishing, include=FALSE, echo=FALSE}
missingData<-dataRaw[dataRaw$num_window %in% difference,]
VariablesNames<-c("num_window","roll_belt","accel_arm_x","accel_arm_y","accel_arm_z","accel_dumbbell_x","accel_dumbbell_y","accel_dumbbell_z","accel_belt_x","accel_belt_y","accel_belt_z","gyros_belt_x","gyros_belt_y","gyros_belt_z","magnet_belt_x","magnet_belt_y", "magnet_arm_x","magnet_arm_y","magnet_arm_z", "gyros_dumbbell_x","gyros_dumbbell_y","gyros_dumbbell_z","magnet_dumbbell_x","magnet_dumbbell_y","magnet_dumbbell_z","pitch_forearm","gyros_forearm_x","gyros_forearm_y","gyros_forearm_z")
Data <- missingData[,VariablesNames]
```

```{r data finishing, include=FALSE, echo=FALSE}

Data$mag_acc_belt<-sqrt((Data$accel_belt_x)^2+(Data$accel_belt_y)^2+(Data$accel_belt_z)^2)

Data$mag_gyro<-sqrt((Data$gyros_belt_x)^2+(Data$gyros_belt_y)^2+(Data$gyros_belt_z)^2)

Data$mag_magnet<-sqrt((Data$magnet_belt_x)^2+(Data$magnet_belt_y)^2)

max_acc_belt<-aggregate(mag_acc_belt~num_window,data=Data,max)

colnames(max_acc_belt)[2]<-"max_acc_belt"

min_acc_belt<-aggregate(mag_acc_belt~num_window,data=Data,min)

colnames(min_acc_belt)[2]<-"min_acc_belt"

var_acc_belt<-aggregate(mag_acc_belt~num_window,data=Data,var)

colnames(var_acc_belt)[2]<-"var_acc_belt"

var_gyro_belt<-aggregate(mag_gyro~num_window,data=Data,var)

colnames(var_gyro_belt)[2]<-"var_gyro_belt"

var_magnet_belt<-aggregate(mag_magnet~num_window,data=Data,var)

colnames(var_magnet_belt)[2]<-"var_magnet_belt"

# Merge the data sets for belt 

relvData<-merge(max_acc_belt,min_acc_belt,by="num_window")

relvData<-merge(relvData,var_acc_belt,by="num_window")

relvData<-merge(relvData,var_gyro_belt,by="num_window")

relvData<-merge(relvData,var_magnet_belt,by="num_window")

# Arm data

Data$mag_accel_arm<-sqrt((Data$magnet_arm_x)^2+(Data$magnet_arm_y)^2+(Data$magnet_arm_z)^2)

max_acc_arm<-aggregate(mag_accel_arm~num_window,data=Data,max)

colnames(max_acc_arm)[2]<-"max_acc_arm"

min_acc_arm<-aggregate(mag_accel_arm~num_window,data=Data,min)

colnames(min_acc_arm)[2]<-"min_acc_arm"

# Merge Arm data

relvData<-merge(relvData,max_acc_arm,by="num_window")

relvData<-merge(relvData,min_acc_arm,by="num_window")

# dumbbell 

Data$mag_gyro_dumbbell<-sqrt((Data$gyros_dumbbell_x)^2+(Data$gyros_dumbbell_y)^2+(Data$gyros_dumbbell_z)^2)

gyro_dumbbell_var<-aggregate(mag_gyro_dumbbell~num_window,data=Data,var)

colnames(gyro_dumbbell_var)[2]<-"gyro_dumbbell_var"

Data$mag_magnet_dumbbell<-sqrt((Data$magnet_dumbbell_x)^2+(Data$magnet_dumbbell_y)^2+(Data$magnet_dumbbell_z)^2)

max_magnet_dumbbell<-aggregate(mag_magnet_dumbbell~num_window,data=Data,max)

colnames(max_magnet_dumbbell)[2]<-"max_magnet_dumbbell"

min_magnet_dumbbell<-aggregate(mag_magnet_dumbbell~num_window,data=Data,min)

colnames(min_magnet_dumbbell)[2]<-"min_magnet_dumbbell"

# Merge dumbbell data

relvData<-merge(relvData,gyro_dumbbell_var,by="num_window")

relvData<-merge(relvData,max_magnet_dumbbell,by="num_window")

relvData<-merge(relvData,min_magnet_dumbbell,by="num_window")

# Glove data

Data$mag_gyros_forearm<-sqrt((Data$gyros_forearm_x)^2+(Data$gyros_forearm_y)^2+(Data$gyros_forearm_z)^2)

max_gyros_forearm<-aggregate(mag_gyros_forearm~num_window,data=Data,max)

colnames(max_gyros_forearm)[2]<-"max_gyros_forearm"

min_gyros_forearm<-aggregate(mag_gyros_forearm~num_window,data=Data,min)

colnames(min_gyros_forearm)[2]<-"min_gyros_forearm"

# Merge glove data

relvData<-merge(relvData,max_gyros_forearm,by="num_window")

relvData<-merge(relvData,min_gyros_forearm,by="num_window")

# "avg_roll_belt

avg_roll_belt<-aggregate(roll_belt~num_window,data=Data,mean)

colnames(avg_roll_belt)[2]<-"avg_roll_belt"

relvData<-merge(relvData,avg_roll_belt,by="num_window")

# "var_roll_belt"

var_roll_belt<-aggregate(roll_belt~num_window,data=Data,var)

colnames(var_roll_belt)[2]<-"var_roll_belt"

relvData<-merge(relvData,var_roll_belt,by="num_window")

# var_accel_arm

Data$mag_accel_arm<-sqrt((Data$accel_arm_x)^2+(Data$accel_arm_y)^2+(Data$accel_arm_z)^2)

var_accel_arm<-aggregate(mag_accel_arm~num_window,data=Data,var)

colnames(var_accel_arm)[2]<-"var_accel_arm"

relvData<-merge(relvData,var_accel_arm,by="num_window")

# "total_accel_dumbbell"

Data$total_accel_dumbbell<-sqrt((Data$accel_dumbbell_x)^2+(Data$accel_dumbbell_y)^2+(Data$accel_dumbbell_z)^2)

total_accel_dumbbell<-aggregate(total_accel_dumbbell~num_window,data=Data,mean)

colnames(total_accel_dumbbell)[2]<-"total_accel_dumbbell"

relvData<-merge(relvData,total_accel_dumbbell,by="num_window")
```

```{r last merge, include=FALSE, echo=FALSE}

# names(testing)
# names(relvData)
# setdiff(names(testing), names(relvData))
# setdiff(names(relvData),names(testing))
total<-rbind(testing,relvData)


```

A new test set, with the name **total**, was produced. 
Predictions on the test set are shown below:

```{r predictions for test set}

predRF_test<-predict(modelRF,newdata=total) 
table(total$num_window, predRF_test)
```

<a name="myfootnote1">1</a>: http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201